---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Edit this file. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# model.form

[![Build](https://github.com/ultreia-io/model.form/actions/workflows/Build.yaml/badge.svg?branch=develop)](https://github.com/ultreia-io/model.form/actions/workflows/Build.yaml?query=branch%3Adevelop)
[![Test](https://github.com/ultreia-io/model.form/actions/workflows/Test.yaml/badge.svg?branch=develop)](https://github.com/ultreia-io/model.form/actions/workflows/Test.yaml?query=branch%3Adevelop)
[![Website](https://img.shields.io/badge/website-documentation-blue)](https://ultreia-io.github.io/model.form/)
[![Coverage reports](https://img.shields.io/badge/coverage-CI_reports-blue)](https://github.com/ultreia-io/model.form/actions/workflows/Test.yaml?query=branch%3Adevelop)
[![Latest release](https://img.shields.io/github/v/release/ultreia-io/model.form?display_name=tag&sort=semver)](https://github.com/ultreia-io/model.form/releases/latest)
[![License](https://img.shields.io/github/license/ultreia-io/model.form)](https://github.com/ultreia-io/model.form/blob/main/LICENSE)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)

Neutral, versioned form definitions with explicit YAML resources and ordered workbook layouts.

**Model checks validate definitions. They do not validate submitted form data.**

## Installation

Download the source archive from [GitHub Releases](https://github.com/ultreia-io/model.form/releases).

```r
install.packages("model.form_0.1.0.tar.gz", repos = NULL, type = "source")
```

Install the runtime dependency `yaml` first if needed. For the development version:

```r
# install.packages("pak")
pak::pak("ultreia-io/model.form@develop")
```

## Start here

```{r}
library(model.form)
directory <- system.file("examples", package = "model.form")
m <- load_form_model("survey", directory)
m
form_field(m, "Observations.unit")[c("enum", "required_field")]
form_layout(m, "Observations")$start_row
check_form_model(form_definition(m))
```

The name `survey` resolves exactly to `survey-form.yaml` in the supplied directory.

Use separate directories for different versions. No directory scanning or latest-version selection occurs.

Explicit file loading and exact version selection remain available:

```{r}
select_form_model(
  "examples/survey-form.yaml", "survey", "2026.1",
  package = "model.form"
)
```

## Contract

* Independent schema and form versions.
* Ordered sheets, fields, fixed cells, and table columns.
* Explicit types and compact constraints, including `mandatory`.
* Declarative references supplied by callers; no embedded domain reference data.
* Structured definition diagnostics and ordinary list/S3 values.
* One runtime dependency: `yaml`. Loading the package has no external side effects.

## Model defaults and conventions

These omitted properties receive values from `model.form`:

| Model YAML                                 | Effective meaning                                                                  |
| ------------------------------------------ | ---------------------------------------------------------------------------------- |
| `mandatory` omitted                        | The field may be absent or null; `form_fields()` returns `mandatory = FALSE`.      |
| Reference `severity` omitted               | An unknown code is an error; `form_fields()` returns `severity = "error"`.         |

Other omissions do not invent values. In particular, an omitted `default` means
that no default is declared. A written `default`, including `default: null`, is
metadata only; `model.form` never inserts it into submitted data.

Every sheet listed under `sheets` is mandatory. There is no sheet-level flag.
Every layout declares `column_count`. For tables, it must equal the number of
fields, whose mapping order defines column order. An optional `row_count` declares
an exact extent. Fixed fields must lie inside all declared dimensions.

`form_definition()` preserves the compact YAML declaration. `form_fields()`
returns effective fields with derived full IDs, columns, and the resolved
defaults above.

## Guides

* [Getting started](https://ultreia-io.github.io/model.form/articles/model.form.html)
* [Model authoring and schema dialect](https://ultreia-io.github.io/model.form/articles/model-authoring.html)
* [Downstream package integration](https://ultreia-io.github.io/model.form/articles/downstream-packages.html)
* [API reference](https://ultreia-io.github.io/model.form/reference/index.html)
* [Contributing](CONTRIBUTING.md) and [maintenance](DEVELOPMENT.md)

## Boundaries

This package loads, checks, and exposes form definitions.

Workbook reading, submitted-data validation, and database loading belong to consuming packages.

## License

LGPL-2.1. See [LICENSE](https://github.com/ultreia-io/model.form/blob/develop/LICENSE).
