---
title: "Distributing models in a downstream package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Distributing models in a downstream package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(model.form)
```


## Package layout

A downstream package owns concrete models. Store each definition as `<name>-form.yaml` under `inst/forms/`.

Use explicit subdirectories when shipping several versions of the same named model.

```text
exampleforms/
  DESCRIPTION
  NAMESPACE
  R/survey.R
  inst/forms/survey-form.yaml
```

Declare `Imports: model.form` in `DESCRIPTION`. Supply a documented wrapper with explicit resource selection.

```r
# In exampleforms/R/survey.R:
survey_model <- function() {
  model.form::select_form_model(
    "forms/survey-form.yaml",
    id = "survey", version = "2026.1",
    package = "exampleforms"
  )
}
```

`package` names the installed package containing the YAML resource.

## Runnable equivalent

The bundled example uses exactly the same installed-resource mechanism.

```{r}
path <- form_resource("examples/survey-form.yaml", "model.form")
stopifnot(file.exists(path))
m <- select_form_model(path, "survey", "2026.1")
form_references(m)
```

Selection reads only supplied paths, validates all of them, and requires exactly one matching identity tuple.

Supply an explicit list of files to support multiple versions. There is no directory scan or latest-version fallback.

Package-relative paths cannot escape the installed package, including through symbolic links.

## Pass models and dependencies explicitly

A future validator accepts a model, submitted data, and caller-supplied reference snapshots.

A form package must not fetch snapshots, connect to a database, or write files during package loading.

Keep concrete domain definitions in the downstream package. Keep loader destinations and mappings in loader packages.

Run model checks in downstream tests for every shipped resource. Reserve validation equivalence tests for the full proof.
