---
title: "Authoring model definitions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Authoring model definitions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(model.form)
```


## Versions and authority

`schema_version` selects this package's contract. Version `1.0.0` is supported by package `0.1.0`.

`form.version` identifies the described model. `schema_version` selects this package's YAML contract.

These are independent strings. Quote versions and codes, including `"01"` and `"2026.1"`.

Comments carry no contract information.

## YAML profile

The only runtime parser is `yaml` (libyaml, YAML 1.1). The package adds structural and semantic checking in R.

Use one UTF-8 document with no document markers, directives, tags, anchors, aliases, or merge keys.

Mappings have unique string keys. Sequences become unnamed lists, including singleton and empty sequences.

Unknown properties are errors at every level. No type coercion is applied.

Omitted-value defaults for `mandatory` and reference severity are described below.

Use lowercase `true` and `false`. YAML 1.1 also parses `yes`, `no`, `on`, and `off` as booleans; quote such codes.

Use decimal integers and finite numbers. YAML 1.1 parses leading-zero integers as octal; quote codes.

Unquoted `null` and `~` become `NULL`. Empty scalars are null, not empty strings. `""` is an empty string.

Only a nonmandatory field's `default` or enum values can be null. Required structural properties cannot be null.

An absent `default` and `default: null` are distinct. Check property presence with `"default" %in% names(field)`.

Dates are ISO strings (`YYYY-MM-DD`); datetimes are UTC strings (`YYYY-MM-DDTHH:MM:SSZ`).

The parser never evaluates YAML expressions, regardless of the `yaml.eval.expr` R option.

## Full example

The following is the actual installed resource, not a second editable copy.

```{r, results='asis', echo=FALSE}
cat("```yaml\n", paste(readLines(form_resource("examples/survey-form.yaml", "model.form")), collapse="\n"), "\n```\n", sep="")
```

## Structural schema

`inst/schema/model-1.0.0.yaml` is the single editable structural schema. `form_schema()` reads the installed file.

The dialect is `model.form/schema/1`, a small schema language; it is deliberately not JSON Schema.

Its interpreter supports these keywords:

| Keyword      | Meaning                                                     |
| ------------ | ----------------------------------------------------------- |
| `ref`        | Resolve a named definition                                  |
| `type`       | mapping, sequence, string, boolean, integer, number, scalar |
| `required`   | Required mapping keys                                       |
| `shorthand`  | Schema for a scalar alternative to a mapping                |
| `properties` | Allowed keys and their schemas; all other keys fail         |
| `items`      | Sequence item schema                                        |
| `min_items`  | Minimum sequence length                                     |
| `min_length` | Minimum string length                                       |
| `pattern`    | Structural string pattern                                   |
| `enum`       | Allowed scalar values                                       |
| `minimum`    | Inclusive numeric lower bound                               |
| `nullable`   | Whether explicit null is allowed                            |

The schema is package-owned, never supplied or overridden by model authors. Unsupported contract versions fail.

This avoids a JavaScript runtime dependency. Semantic invariants remain explicit R checks with behavioural tests.

```{r}
schema <- form_schema()
names(schema$definitions)
schema$definitions$field
```

## Identifiers and field meaning

Identifiers start with an ASCII letter and then contain letters, digits, dots, underscores, or hyphens.

Sheet names and local field names are mapping keys. Full field IDs are derived as `<sheet>.<field>`.

Field types are `string`, `integer`, `number`, `boolean`, `date`, and `datetime`.

An enum uses `enum` on one of these types; see [Enumerated fields](#enumerated-fields).

`mandatory: true` requires a filled, non-null value. Omission means the field may be empty.

A `default` is descriptive metadata; this package never fills absent submitted values.

`type` is explicit for every field. Write `mandatory` first when true, then cell `location`, `type`, and other rules.

## Omitted values and effective fields

The compact YAML declaration omits common values. The documented omitted properties
are resolved by `model.form`:

| Property                             | When omitted                      | Effective accessor value |
| ------------------------------------ | --------------------------------- | ------------------------ |
| `mandatory`                          | The field may be absent or null   | `FALSE`                  |
| `severity` on a reference constraint | An unknown code is an error       | `"error"`                |

For example, these declarations are equivalent when read through `form_fields()`:

```yaml
PORT_CODE: {type: string, reference: refs_admin.ports}
COMMENT: {type: string}
```

```yaml
PORT_CODE: {mandatory: false, type: string, reference: {id: refs_admin.ports, severity: error}}
COMMENT: {mandatory: false, type: string}
```

`form_definition()` returns the compact declaration and therefore preserves
omission. `form_fields()` returns effective field definitions: it adds
`mandatory = FALSE`, expands reference IDs with severity `"error"`, derives the full
`<sheet>.<field>` ID, and derives table columns from field order.

No other optional property receives a value. In particular:

* omitted `default` means no default metadata is declared;
* `default: null` explicitly declares null metadata;
* omitted validation rules add no checks;
* omitted `row_count` means no exact row extent is declared.

A declared `default` never fills an absent submitted value. It is metadata for
consumers. Validation and data-loading packages decide whether they use it.

Every sheet present in the `sheets` mapping is mandatory. This rule is inherent
in the model contract; there is no omitted sheet property being defaulted.

## Declarative constraints

* `mandatory`: requires a filled value; omitted means `false`.
* `enum`: nonempty list of distinct, typed values; nonmandatory fields may include null.
* `minimum`, `maximum`: inclusive bounds for numbers and integers; integer bounds must be whole numbers.
* `pattern`: a valid PCRE regular expression for string, date, or datetime fields.
* `unique`: declares uniqueness within the field's repeated table values.
* `reference`: a declared ID (`reference: stations`) defaults to `error`; use `{id: stations, severity: warning}` for warnings.
* `required_field`: require a value when the named local field is filled; see [Conditional fields](#conditional-fields).

The required field must be on the same sheet and cannot be the field itself. Cycles are allowed: no rules execute here.

References are names of caller-supplied string code lists.

They contain neither database locations nor reference data. The caller supplies snapshots to downstream validation.

Defaults and enum values must match the field type, mandatory policy, numeric bounds, and pattern.

Defaults must also belong to an enum when one is declared. References are checked structurally, without fetching data.

Blank-cell handling, severity for other rules, and execution belong to `model.validation`.

### Conditional fields

`required_field` names one companion field on the same sheet:

```yaml
unit: {type: string, enum: [cm, m], required_field: length}
```

The unit is required when `length` is filled, including when its value is zero.

For a table, the dependency is evaluated in the same row. The property accepts one name, never a list or matching mode.

This condition does not forbid a value when the companion field is empty.

### Enumerated fields

`enum` restricts the allowed values of a field; it is not a separate field type. Do not write `type: enum`.

For a string field, use this field fragment:

```yaml
type: string
default: cm
enum: [cm, m]
```

For an integer field:

```yaml
type: integer
default: 2
enum: [1, 2, 3]
```

The list must be nonempty, contain no duplicates, and use values compatible with the declared field type.

A declared default must belong to the enum. Include `null` only for a nonmandatory field.

Quote numeric-looking string codes, for example `enum: ["01", "02"]` on a `string` field.

These checks validate the declaration. Checking submitted values against it belongs to downstream validation.

## Workbook layouts

`sheets` is an ordered mapping keyed by physical worksheet name. Every declared sheet is mandatory.

Sheet names must be unique ignoring case. Declaration order records expected sheet order.

Each sheet's `fields` mapping is keyed by local field name.

Put each sheet's `layout` before its `fields`.

Cell fields require an uppercase A1 address, such as `location: E3` or `location: AA12`.

```yaml
META:
  layout: {kind: cells, row_count: 12, column_count: 5}
  fields:
    form_version: {mandatory: true, location: E3, type: string, pattern: '^1\.0\.0$'}
```

`form_fields()` derives numeric `row` and `column` coordinates. Raw declarations retain only `location`.

For a table, field order is column order; fields must omit `location`.

Table layouts require `start_row` and `column_count`. The column count must equal
the number of declared fields. Rows before `start_row` are preamble; their text is not part of the definition.

Cell layouts require `column_count`. Every fixed field must lie within declared
dimensions. All coordinates are one-based integers.

An optional `row_count` declares an exact physical row extent. Empty table
sheets may use `column_count: 0` when they declare no fields.

Field locations must lie within declared dimensions. Empty tables may start immediately after the last declared row.

Multiple fields cannot occupy one location. Fields retain their declaration order.

An empty placeholder sheet uses `fields: {}`.

## Evolution

Changes to model meaning should change `form.version`.

Schema evolution requires an explicit new contract version; unknown properties do not act as extension points.

## Diagnostics

`check_form_model()` returns `valid` and a diagnostic data frame with `resource`, `path`, and `message` columns.

Property paths start at `$` and use mapping keys or one-based sequence indices.

Structural errors are collected first. Semantic checks run only after structural success.
