---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(model.form)
```


The package turns a YAML form description into a checked R list, then provides functions to inspect it.

It describes fields, constraints, and workbook layouts. It does not read or validate completed workbooks.

To try the examples from a project checkout, first run this in the project directory:

```r
pkgload::load_all()
```

For an installed package, use `library(model.form)` instead.

## Load an installed example

```{r}
directory <- system.file("examples", package = "model.form")
m <- load_form_model("survey", directory)
m
```

The example ships under `inst/examples/`, so it remains available after installation.

The name `survey` resolves to `survey-form.yaml` in that directory. This is the standard naming convention.

Choose the version directory explicitly; the loader never searches for a latest version.

`read_form_model(path, package)` remains available when you need to load an explicit filename.

## What a model contains

A `form_model` is the checked definition itself, with an S3 class:

```{r}
names(m)
```

It directly contains the form identity, versions, references, and sheets.

Sheets use worksheet names, and their fields use local field names:

```{r}
names(m$sheets)
names(m$sheets$Metadata$fields)
m$sheets$Metadata$fields$submission[c("mandatory", "location", "type")]
```

The YAML path is used in loading errors, then discarded after a successful load.

The S3 class selects the compact `print()` method shown above. It does not make the object immutable.

## How the R code fits together

The files organize functions in one package namespace. You do not source each file separately.

| File        | Responsibility                                          |
| ----------- | ------------------------------------------------------- |
| `read.R`    | Locate resources and parse the restricted YAML format    |
| `check.R`   | Check definitions and construct valid model objects      |
| `inspect.R` | Retrieve definition parts and print summaries            |

Loading follows this call sequence:

```text
read_form_model()
  +-- form_resource()       Resolve the path when package is supplied
  +-- read_yaml_data()
  |     +-- yaml_profile()  Reject excluded YAML constructs
  |     +-- yaml::yaml.load() Parse data without evaluating expressions
  +-- as_form_model()
        +-- check_form_model()
              +-- form_schema()    Load the package-owned structural schema
              +-- schema_check()   Check properties and value types
              +-- semantic_check() Check references and consistency
```

For example, structural checking rejects a field without a `type` or a string where a number is required.

Semantic checking rejects duplicate cell coordinates or a numeric minimum greater than its maximum.

Semantic checks run only after structural checks pass. This avoids follow-on errors caused by malformed definitions.

`check_form_model()` returns findings; `as_form_model()` raises an error if they prevent constructing a valid model.

Functions such as `yaml_profile()` and `schema_check()` are internal helpers, not entry points for callers.

For code review, start with `read_form_model()`, then `as_form_model()`, `check_form_model()`, and `form_field()`.

## Inspect fields and locations

```{r}
names(form_fields(m, "Observations"))
field <- form_field(m, "Observations.unit")
field$type
field$enum
layout <- form_layout(m, "Observations")
layout$start_row
form_field(m, "Metadata.submission")[c("row", "column")]
```

`form_field()` finds a field by its exact ID across all sheets. Collection accessors name their entries by ID.

`form_layout()` uses the physical worksheet name.

Lists retain declaration order. Locations use one-based row and column numbers.

Accessors return declarations; they do not execute constraints or recheck the model.

For example, downstream validation must implement the declared condition and supply reference data:

```{r}
field$required_field
form_field(m, "Observations.station")$reference
form_references(m)
```

`form_references()` describes required datasets; it neither downloads them nor connects to a database.

## Diagnose a definition

```{r}
d <- form_definition(m)
d$sheets[[2]]$fields[[1]]$reference <- "missing"
result <- check_form_model(d, resource = "my-definition.yaml")
result$diagnostics
stopifnot(!result$valid)
```

`as_form_model()` and `read_form_model()` throw `model_form_error` when a definition is invalid.

```{r}
error <- tryCatch(as_form_model(d, "my-definition.yaml"), model_form_error = identity)
error$path
error$diagnostics
```

## Edit a copy

```{r}
d <- form_definition(m)
d$form$version <- "2026.2"
copy <- as_form_model(d, "edited in memory")
stopifnot(form_metadata(m)$form$version != form_metadata(copy)$form$version)
```

Editing the extracted definition leaves `m` unchanged. `as_form_model()` checks the edited copy before returning it.

Directly changing `m` bypasses those checks. Reconstruct the model after edits.

No reference data, credentials, workbook readers, or live services are needed for these examples.
